---
title: "Sample Descriptives"
format: html
---

The first artifact in your manuscript will probably be a table of sample descriptives.

## Using the tableone package

This example shows an end-to-end workflow: recoding numeric-coded demographic variables into labeled factors, building a "Table 1"-style summary table with **tableone**, converting it into a formatted **flextable** object, and exporting the result directly to a Word document.

The **tableone** package is great because it's whole purpose is literally this, so no other functionality is needed. However, it does not allow exporting directly to a Word document, so I have provided a function to do this that some other fantastic, altruistic genius created and made freely available.

**Packages needed:**

-   dplyr (or tidyverse) [@tidyverse2019]

-   tableone [@tableone]

-   flextable [@flextable]

-   officer [@officer]

### Step 1: Recode numeric codes into labelled factors

Let's generate some fake data.

```{r}
library(dplyr)

set.seed(1234)

data <- data.frame(
  age = round(rnorm(500, mean = 42, sd = 15)),
  gender = sample(
    c(1, 2, 3, 4),
    size = 500,
    replace = TRUE,
    prob = c(.47, .49, .02, .02)
  ),
  state = sample(
    1:8,
    size = 500,
    replace = TRUE,
    prob = c(.31, .25, .20, .08, .10, .03, .01, .02)
  ),
  education = sample(
    1:5,
    size = 500,
    replace = TRUE,
    prob = c(.08, .25, .20, .30, .17)
  )
)

# Keep age within a reasonable range
data$age <- pmin(pmax(data$age, 18), 85)
```

Now let's recode.

```{r}
data <- data %>%   
  mutate(gender_r = factor(gender,                           
                           levels = c(1, 2, 3, 4),                           
                           labels = c("Male", "Female", "Non-binary", "Other")),
         state_r = factor(state,
                          levels = c(1, 2, 3, 4, 5, 6, 7, 8),
                          labels = c("NSW", "VIC", "QLD", "SA", "WA", "TAS", "NT", "ACT")),
         education_r = factor(education,
                              levels = c(1, 2, 3, 4, 5),
                              labels = c("Less than high school", "High school", "TAFE", "Undergraduate degree", "Postgraduate degree")))
```

**What this is doing:**

-   This builds on the [factor-conversion approach](recoding.qmd), but adds a `labels =` argument alongside `levels =`. `levels` tells R the *order* and *original numeric values* to match against; `labels` tells it what *text* to display in place of each number — so `1` becomes `"Male"`, `2` becomes `"Female"`, and so on.

### Step 2: Build the summary table with createTableOne()

```{r}
library(tableone)  

vars <- c("age", "gender_r", "education_r")  # Select variables
table_1 <- CreateTableOne(data = data,
                          vars = vars,
                          strata = "state_r",
                          addOverall = TRUE,
                          test = FALSE) 

table_1
```

**What this is doing:**

-   `vars <- c("age", "gender_r", "education_r")` defines which variables should appear as rows in the table..

-   `CreateTableOne()` builds the summary table itself. It automatically detects each variable's type and picks an appropriate summary: means and SDs for continuous variables like `age`, counts and percentages for factors like `gender_r`and `education_r`— you don't need to specify this manually.

-   `strata = "state_r"` splits the table into separate columns, one per state, so you can see how demographics differ by state at a glance, rather than only getting one pooled summary. This is useful if you have a representative survey; however, you could also list this as another count variable like gender and education. Similarly, you could add in an ethnicity variable, if relevant.

-   `addOverall = TRUE` adds an additional column showing the summary across *all* participants combined, alongside the state-by-state breakdown. This is useful if you have stratified the table, as we have done here.

-   `test = FALSE` turns off tableone's default behavior of automatically running significance tests (e.g. ANOVA, chi-square) comparing groups and adding a p-value column. Turn this off if the table is meant to be purely descriptive (as Table 1s usually are). Turn this on with `TRUE` if you specifically want group comparisons included.

-   Printing `table_1` on its own displays the table in your R console.

### Step 3: Converting the tableone object into a flextable

This is a function helpfully created by Graham Cooper [@Cooper2024] and made available on GitHub to quickly convert a **tableone** object to something that you can open in a Word document using the **flextable** and **officer** packages. Make sure you have those two packages installed and loaded to be able to run the function.

```{r}
tableone2flextable <- function(tableone) {

  tableone <- print(
    tableone,
    quote = FALSE,
    noSpaces = TRUE
  )

  rows <- nrow(tableone)
  cols <- ncol(tableone)
  column_names <- colnames(tableone)

  listoflists <- vector("list", cols)

  for (i in seq_len(cols)) {

    start <- (i - 1) * rows + 1
    end <- i * rows

    listoflists[[i]] <- tableone[start:end]
  }

  dataframe <- as.data.frame(
    listoflists,
    col.names = column_names
  )

  dataframe <- cbind(
    Variable = rownames(tableone),
    dataframe
  )

  rownames(dataframe) <- NULL

  flextable::flextable(dataframe)
}
```

**What this is doing:**

-   `print(tableone, quote = FALSE, noSpaces = TRUE)` converts the tableone object into a plain **character matrix**, stripping out quote marks and extra padding spaces that tableone normally adds for console display.

-   `nrow()`, `ncol()`, and `colnames()` capture the matrix's dimensions and column headers, which are needed to correctly rebuild the table as a dataframe in the next steps.

-   The `for` loop walks through the matrix column by column (`1:cols`), extracting each column's values from the matrix (which is stored internally as one long vector) using `start`/`end` index positions, and stores each column as one element of `listoflists`.

-   `as.data.frame(listoflists, col.names = colnames)` combines those extracted columns into an actual dataframe, restoring the original column headers.

-   `cbind(Variable = rownames(tableone), dataframe)` adds the row labels (variable names like "age," "gender_r") back in as their own proper column, since they'd otherwise just be dataframe row names rather than visible data.

-   `rownames(dataframe) <- NULL` resets the dataframe's row names back to plain sequential numbers, since the row names were just used above and aren't needed anymore as actual row labels.

-   `flextable::flextable(dataframe)` converts the finished dataframe into a **flextable** object — the format needed for the polished formatting and Word export in the next step. Using `flextable::` here (rather than `library(flextable)` beforehand) explicitly specifies which package the function comes from, without needing to load the whole package into your session.

### Step 4: Format and export to Word

```{r}
flextable2word <- function(flextab, filepath) {

  ft <- flextable::theme_apa(flextab) %>%
    flextable::autofit() %>%
    flextable::align(
      j = 1,
      align = "left",
      part = "all"
    ) %>%
    flextable::line_spacing(
      space = 1,
      part = "all"
    )

  officer::read_docx() %>%
    flextable::body_add_flextable(ft) %>%
    officer::body_end_section_landscape() %>%
    print(target = filepath)

  browseURL(filepath)
}

tab_1 <- tableone2flextable(table_1)
tab_1
```

Save and open Word document.

**What this is doing:**

-   `flextable2word` is a second custom function that takes a flextable object (from Step 3) and a `filepath` argument, letting you specify exactly where and under what filename the finished Word document should be saved.

-   `require(tidyverse)` and `require(flextable)` ensure both packages are loaded whenever this function runs, even if you forgot to load them earlier in your script.

-   `theme_apa(flextab)` applies flextable's built-in **APA-style formatting** preset.

-   `autofit(...)` automatically adjusts column widths to fit their content, rather than leaving default widths that might cut off longer text.

-   `align(j = 1, align = "left", part = "all")` left-aligns the first column (`j = 1`, i.e. your `Variable` column) across the whole table (`part = "all"` covers header and body).

-   `line_spacing(space = 1, part = "all")` sets single line spacing throughout the table, overriding Word's common default of double spacing, which looks cluttered in a table context.

-   `officer::read_docx()` starts a new, blank Word document object using the **officer** package (which flextable relies on for Word export) — `officer::` here again explicitly specifies the package without a separate `library()` call.

-   `body_add_flextable(ft)` inserts your formatted table into that blank document.

-   `body_end_section_landscape()` sets the page orientation to landscape for this section — useful for wider tables (like this one, with multiple state columns) that wouldn't fit well on a standard portrait page.

-   `print(target = filepath)` writes the finished Word document out to disk, at the location and filename you specify in `filepath`.

-   `browseURL(filepath)` automatically opens the newly created Word file once it's saved, so you can immediately check the result without manually navigating to find it.

### Step 5: Run it

``` r
flextable2word(
  tab_1,
  "output/Table 1 Sample demographics.docx"
)
```

**What this is doing:**

-   `tableone2flextable(table_1)` runs the Step 3 function on the table you built in Step 2, producing the flextable version.

-   `flextable2word(tab_1, "output/Table 1 Sample demographics.docx")` runs the Step 4 function, exporting the finished, formatted table straight to a Word file in your `output/` folder.

-   Running `tab_1` on its own at the end displays the flextable directly in your R/RStudio viewer, letting you do a final visual check.
